HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-04-24 docs/HLL_exp_bad/main.tex
more work

\section{Vulnerabilities}\label{sec:attacks}
This section presents different adversarial settings and the corresponding attacks to manipulate the HLL estimate. The attacker's goal in our work is to bring the estimate down, as this would cause the most damage when looking at the applications of HLL. One can think about the attacker trying the hide from a detection system. Conveniently, it is easy to adopt the same strategy as the one we present for an attacker who is looking to artificially inflate the HLL estimate given a limited data set input. The impact over the cardinality estimate and the computational effort to be done by the attacker is analyzed afterwards.

\subsection{Adversarial setting}
We model 3 adversarial scenarios based on the attacker's knowledge and capabilities. For all of them, we assume that the sketch already contains some existing data from honest users to simulate a real life setup. Among the scenarios, S3 has the strongest assumptions as it combines the knowledge of attackers from S1 and S2.

S1: the attacker is able to test insertion of an item on a black-box snapshot of the sketch in use and check the effect on the cardinality estimate.

S2: the attacker has access to the details of the implementation, i.e., the number of buckets and the hash function $h$.

S3: the attacker has direct access to the sketch in use and all internals, i.e., the number of buckets, the hash function $h$, and also the sketch's buckets content. Those buckets hold the intermediate estimators updated by the honest user's data, that are averaged in a later step to compute the global cardinality estimator.\\

S1 and S2 are presented as M2 and M1 respectively in \cite{hllvuln}. We do not put our focus on S1 as this previous work fully exploit this scenario. The attack works as follow: ``after each insertion the cardinality is checked and the items which do not increment the HLL estimate are retained``. We start by presenting our attack in S2, providing much stronger results then the corresponding attack of \cite{hllvuln}, and then use the additional information the attacker has access to in S3 to make the attack even more efficient.

\subsection{Manipulating the estimate}
Under S2: The attacker has access to $h$ so he can adjust some field of the item $x$ he is trying to insert and keep only the one such that the $l$ rightmost bits of $h(x)$ start with a 1. Indeed, the number of leading 0 bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the averaged approximated cardinality is left unchanged. As the attacker knows how many buckets are used, he can recover $n$ and $l$, and aim for the $n$th bit (first bit not used for bucket allocation) to be 1. For example, in the scenario of \cite{portscanhll}, the attacker could change the source port of his IP packet $x$ until he is satisfied with $h(x)$. \textbf{(what about repeated sampling?)}\\

Under S3: The attacker can adopt the same strategy but make the attack more efficient and reduce the sampling requirements with the additional information. Now that the attacker has the sketch, not only he knows the overall estimated cardinality but also the intermediate estimators computed per bucket. So, he can "allow" some leading 0 bits in the $l$ rightmost bits of $h(x)$ as long as it does not increment the bucket's estimate. In other words, if $h(x)$ is mapped to the $i$th bucket with estimated cardinality $2^{c_i + 1}$ (it observed a value with $c_i$ leading 0s), then, instead of the strict restriction of a 1 bit at the $n$th index, the attacker is satisfied when having at least one 1 bit in the $n$ to $n+c_i + 1$ bit slice of $h(x)$. In case the attacker hits an empty bucket (the estimator is 0), he must skip that bucket completely. Under those conditions, the estimator of the buckets are not updated, and there is no effect on the cardinality estimation of the HLL sketch.

@@

Under S2, the attacker need to craft 2 $x$s on average to obtain a 1 bit at the $n$th position of $h(x)$ and successfully perform the attack. As we do not care about the remaining bits, the attacker can theoretically "hide" up to 2$^{len(h(x))-1}$ items into the sketch.\\

Under S3, let's compute the average work for the attacker to find an $x$ such that $h(x)$ meets the attack's requirements. We assume $h(x)$ is mapped to bucket $i$. The probability that $h(x)$ does not increment the estimator is the probability that $h(x)$ has less than $c_i$ leading 0s. So:\\
$P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i)=P(h(x)$ has LESS than $c_i$ leading 0s$)=1-P(h(x)$ has strictly MORE than $c_i$ leading 0s$)=1-\frac{1}{2^{c_i+1}}$\\
Across all buckets, the attacker does not update any estimate with probability $P=
\sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i$ AND $h(x)$ does NOT update bucket $i) = \sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i)P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i) = \sum_{i=1}^{2^n}\frac{1}{2^n} \times (1-\frac{1}{2^{c_i+1}}) = 1-\frac{1}{2}\times(\frac{1}{2^n}\times \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) = 1-\frac{1}{2}\times\frac{1}{H_c} = 1-(2H_c)^{-1}$, with $H_c$ the harmonic mean of the numbers $2^{c_1}, 2^{c_2}, ..., 2^{c_{2^n}}$.
So the attacker needs to craft on average $(1-(2H_c)^{-1})^{-1} x$s, and can theoretically "hide" up to 2$^{len(h(x))}\times (1-(2H_c)^{-1})$ items into the sketch.

@@

We assume here that there already are values coming from other honest users' data in the sketch, we now remove this assumption. Unfortunately, the attacker in S2 has no way to detect that he is attacking an empty sketch and cannot adapt his strategy, so all buckets hit will update once their estimator to $2^{0+1}=2$ items, leading to the sketch final approximation of the number of distinct elements to $\alpha2^{n+1}$ regardless of the amount of items the attacker will send. However, the attacker in S3 is able to notice that all buckets are empty and adapt his attack. As mentioned above, the attacker in S3 avoids empty buckets, but when dealing with an empty sketch, sacrifices must be made if the attacker wants to insert at least one item. To limit the detection, the attacker can target and fill one bucket only. This would fix $n+1$ bits of $h(x)$, allowing the attacker to add $2^{len(h(x))-n-1}$ elements while keeping the cardinality to 1.

@@

We challenge the attacker to pick the largest possible number of items to add to the sketch while trying to keep the cardinality estimate as low as possible, either in the S2 or the S3 scenario. In a second time, in order to get results that can be compared to the work of \cite{hllvuln}, we mount our attack in their more restrictive setting: instead of picking from the universe of possible strings, the attacker has to choose from a set of 250'000 random items that are given to him. Finally , for completeness, we also run the attacks against empty sketches. All attacks are ran 30 times and the outcomes are reported in Tab. \ref{table:tab1} and Tab. \ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the attacker and the approximation of the cardinality after adding the attacker's items.

Those empirical results confirm our impact analysis on several points. First, when attacking a sketch already containing some items, we can see that the attacker is able to hide approximately half of the items' set under S2, and around 83\%%( $205$'$000 \approx 250'000 \times 0.82 \approx 250$'$000\times(1-\frac{1}{2\times H_c})$, with $H_c\approx\frac{1'000}{256}=3.9$).
$\approx (1-\frac{1}{2\times H_c})$ with $H_c\approx\frac{1'000}{256}=3.9$, under S3.
Secondly, when attacking an empty sketch, the final cardinality estimate is always exactly 367 $=\alpha2^{n+1}$ with $\alpha = 0.72$ as specified by the Eq. (29) of \cite{hll2} or 1 in respectively S2 and S3.

To assess the efficiency of our attacks, we can confront our results from Tab. \ref{table:tab2} with the attack presented in the Section 5.1 of \cite{hllvuln}, where they target the Redis \cite{redishll} open-source implementation of HLL and perform the attack against an empty sketch. Given a set of 250'000 items, the attacker could choose ``74'390 distinct items and obtain an HLL estimate of only 15'780`` achieving a five-fold reduction from the true cardinality. In comparison, the attacker following our strategy is able to add 125'000 distinct items while keeping the estimate at 367, demonstrating a reduction by a factor of 340.

>>>>

\section{HLL in Adversarial Settings}\label{sec:attacks}

This section presents different adversarial settings and corresponding attacks that manipulate the HLL cardinality estimate. The adversary's goal is to reduce the estimate as much as possible, since this would cause more damage when  considering the above-mentioned applications of HLL. It is easy to modify our attack strategies for an adversary who instead wishes to artificially inflate the HLL estimate given a limited input capability. We also analyze the impact on the HLL cardinality estimate and the computational effort required.

Throughout, we assume the adversary has the ability to vary the contents of free fields in the input strings which will be inserted into the stream of values. For example it can manipulate IPID and other header fields in the context of IP packets. This enables the adversary to control the output of hash function $h$ and flexibly choose which items will be inserted into the HLL, and in which buckets.

\subsection{Adversarial Setting}
We model three distinct adversarial scenarios based on the adversary's knowledge and capabilities. For all of them, we assume that the sketch already contains some existing data from honest users to simulate a real-life setting. Among the scenarios, S3 makes the strongest assumptions about the adversarial capabilities, since it combines the knowledge of adversarys from S1 and S2.

\begin{itemize}
\item[S1:] the adversary has black-box access to the HLL in use -- that is, the adversary can ask for the HLL's cardinality estimate at any point in time, and so is able to check the effect on the cardinality estimate when inserting an item; 
\item[S2:] the adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use.
\item[S3:] the adversary has direct access to the sketch in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL. Recall that these buckets hold the intermediate estimators that are averaged in a later step to compute the final cardinality estimate.
\end{itemize}

Scenarios S1 and S2 are presented as M2 and M1, respectively, in~\cite{hllvuln}. We do not put our focus on S1 as this previous work in~\cite{hllvuln} fully exploits this scenario. Their attack works as follow: ``after each insertion the cardinality is checked and the items which do not increment the HLL estimate are retained''. It is unclear what this means in practice, since after insertion into the HLL, an item cannot usually be removed by the adversary. \kenny{Matilda, can you think about this sentence that I added? Do we need to rewrite S1? The scenario does not really make sense at all in practice. I think we might use this to beat up the other paper a bit more.}

We start by presenting our attack in scenario S2, providing much stronger results than the corresponding attack of \cite{hllvuln}, and then use the additional information the adversary has access to in scenario S3 to make the attack even more powerful.

\subsection{Manipulating the HLL Cardinality Estimate}

\paragraph{Scenario S2:} The adversary knows $h$ so it can adjust fields in the item $x$ it is trying to insert and keep only those for which the $\ell$ rightmost bits of $h(x)$ start with a one bit. In this case, the number of leading zero bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the averaged approximate cardinality is likely to be left unchanged. For example, in the scenario of \cite{portscanhll}, the adversary could change the source port of his IP packet until it is satisfied with its hash value. Given sufficient flexibility in the input $x$, the adversary can insert many items into the HLL without increasing the cardinality markedly, by repeatedly sampling from the input domain.

\paragraph{Scenario S3:} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the intermediate estimators computed per bucket). So, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increment the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.  A special case arises when the HLL is completely empty at the beginning of the attack; here the adversary is forced to increase the estimate in at least one bucket, but with sampling this can obviously be done in such a way as to increase the HLL estimate to $2$ in a single bucket, leaving all the other buckets untouched.

@@

Under S2, the adversary need to craft 2 $x$s on average to obtain a 1 bit at the $n$th position of $h(x)$ and successfully perform the attack. As we do not care about the remaining bits, the adversary can theoretically "hide" up to 2$^{len(h(x))-1}$ items into the sketch.\\

Under S3, let's compute the average work for the adversary to find an $x$ such that $h(x)$ meets the attack's requirements. We assume $h(x)$ is mapped to bucket $i$. The probability that $h(x)$ does not increment the estimator is the probability that $h(x)$ has less than $c_i$ leading 0s. So:\\
$P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i)=P(h(x)$ has LESS than $c_i$ leading 0s$)=1-P(h(x)$ has strictly MORE than $c_i$ leading 0s$)=1-\frac{1}{2^{c_i+1}}$\\
Across all buckets, the adversary does not update any estimate with probability $P=
\sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i$ AND $h(x)$ does NOT update bucket $i) = \sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i)P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i) = \sum_{i=1}^{2^n}\frac{1}{2^n} \times (1-\frac{1}{2^{c_i+1}}) = 1-\frac{1}{2}\times(\frac{1}{2^n}\times \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) = 1-\frac{1}{2}\times\frac{1}{H_c} = 1-(2H_c)^{-1}$, with $H_c$ the harmonic mean of the numbers $2^{c_1}, 2^{c_2}, ..., 2^{c_{2^n}}$.
So the adversary needs to craft on average $(1-(2H_c)^{-1})^{-1} x$s, and can theoretically "hide" up to 2$^{len(h(x))}\times (1-(2H_c)^{-1})$ items into the sketch.

@@

We assume here that there already are values coming from other honest users' data in the sketch, we now remove this assumption. Unfortunately, the adversary in S2 has no way to detect that he is attacking an empty sketch and cannot adapt his strategy, so all buckets hit will update once their estimator to $2^{0+1}=2$ items, leading to the sketch final approximation of the number of distinct elements to $\alpha2^{n+1}$ regardless of the amount of items the adversary will send. However, the adversary in S3 is able to notice that all buckets are empty and adapt his attack. As mentioned above, the adversary in S3 avoids empty buckets, but when dealing with an empty sketch, sacrifices must be made if the adversary wants to insert at least one item. To limit the detection, the adversary can target and fill one bucket only. This would fix $n+1$ bits of $h(x)$, allowing the adversary to add $2^{len(h(x))-n-1}$ elements while keeping the cardinality to 1.

@@

We challenge the adversary to pick the largest possible number of items to add to the sketch while trying to keep the cardinality estimate as low as possible, either in the S2 or the S3 scenario. In a second time, in order to get results that can be compared to the work of \cite{hllvuln}, we mount our attack in their more restrictive setting: instead of picking from the universe of possible strings, the adversary has to choose from a set of 250'000 random items that are given to him. Finally , for completeness, we also run the attacks against empty sketches. All attacks are ran 30 times and the outcomes are reported in Tab. \ref{table:tab1} and Tab. \ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary and the approximation of the cardinality after adding the adversary's items.

Those empirical results confirm our impact analysis on several points. First, when attacking a sketch already containing some items, we can see that the adversary is able to hide approximately half of the items' set under S2, and around 83\%%( $205$'$000 \approx 250'000 \times 0.82 \approx 250$'$000\times(1-\frac{1}{2\times H_c})$, with $H_c\approx\frac{1'000}{256}=3.9$).
$\approx (1-\frac{1}{2\times H_c})$ with $H_c\approx\frac{1'000}{256}=3.9$, under S3.
Secondly, when attacking an empty sketch, the final cardinality estimate is always exactly 367 $=\alpha2^{n+1}$ with $\alpha = 0.72$ as specified by the Eq. (29) of \cite{hll2} or 1 in respectively S2 and S3.

To assess the efficiency of our attacks, we can confront our results from Tab. \ref{table:tab2} with the attack presented in the Section 5.1 of \cite{hllvuln}, where they target the Redis \cite{redishll} open-source implementation of HLL and perform the attack against an empty sketch. Given a set of 250'000 items, the adversary could choose ``74'390 distinct items and obtain an HLL estimate of only 15'780`` achieving a five-fold reduction from the true cardinality. In comparison, the adversary following our strategy is able to add 125'000 distinct items while keeping the estimate at 367, demonstrating a reduction by a factor of 340.
